A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-07 paper/paper.tex
clarity edits to intro

\begin{abstract}
  We present a domain-specific type theory for constructions and
  proofs in category theory. The type theory axiomatizes notions of
  category, functor, profunctor and a generalized form of natural
  transformations. By restricting the type theory to an ordered linear
  restriction of predicate logic, we guarantee that all functions
  between categories are functorial, all relations are profunctorial,
  and all transformations are natural by construction, with no
  separate proof necessary. Important category theoretic proofs such
  as the Yoneda lemma and Co-yoneda lemma become simple type theoretic
  proofs about the relationship between unit, tensor and (ordered)
  function types, and can be seen to be ordered refinements of
  theorems in predicate logic.

@@

Category theory is a branch of mathematics that gives a systematic
theory of typed, ``higher-dimensional'' algebraic structures.
%
Originally developed for applications to homological algebra, it was
quickly discovered that (among many other applications) these
categorical structures were common in logic and computer science, and
that formal systems like logics, type theories and programming
languages can be modeled by sound and complete semantic models given
by notions of structured category\cite{lawvere,goguen,lambek,moggi}, a
phenomenon sometimes referred to as the Curry-Howard-Lambek
nncorrespondence \cite{curry-howard-lambek}.
%
Examples include the simply typed lambda calculus \cite{lambekscott},
computational lambda calculus \cite{moggi}, linear logic \cite{girard}
and dependent type theory\cite{not-sure}.
%
These seminal works have led to a fruitful field of research known as
categorical logic, where new type theories are designed based directly
on category-theoretic semantics.
%
The slogan is that the syntax of a type theory should present an
initial object in its category of models, a category-theoretic
reformulation and generalization of the logical notion of soundness
and completeness.

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex, and there is difficulty in
acquisition of competence in category theory by computer scientists
who do not have a strong background in pure mathemantics.
%
While much of this can be explained by lack of, or lack of awareness
of, accessible material for computer scientists, in some ways category
theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and Kan extensions produce a proliferation of ``naturality''
and ``functoriality'' side-conditions that must be discharged.
%
This is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
These generalizations have long been recognized as being important in
programming language applications: domain-,metric- and
step-index-enriched categories have been used to model general
recursion and internal categories have been used to model
parametricity and gradual
typing\cite{order-enriched-cats,topos-of-trees,reflexive-graphs,double-cats-gradual-typing},
to name a few.

% TODO: Yoneda example
Novices in category theory are often overwhelmed by the number of
naturality and functoriality side conditions that abound in otherwise
simple proofs in category theory. For instance consider the proof of
the Yoneda lemma, often the first ``big'' proof in an introductory
category theory text. The Yoneda lemma states that for any functor $P
: C^o \to \Set$ and object $A \in C$, that the set of natural
transformations $C(-,A) \to P$ is in bijection with the elements
$P(A)$. Furthermore, this bijection is natural in both $A$ and $P$. In
the authors' experience teaching category theory the sheer amount of
naturality side conditions in this result can be very intimidating to
learners, while experienced practitioners will observe that these
naturality conditions are tedious to carry out but routine. In our
type theory, none of these naturality conditions need to be explicitly
proven. Instead, all terms expressible in the type theory are
automatically natural by construction.

Fortunately, the tools of category theory itself can be employed to
simplify these constructions, specifically the tools of \emph{higher}
category theory.
%
Consider the analogous situation in differential calculus. When an
adept analyst writes down a function, they do not expand out the
$\epsilon-\delta$ definition of continuity and prove that it is
continuous. Rather, they are aware of certain \emph{syntactic
principles}, most importantly that composition of continuous functions
is continuous and that previously defined functions are known to be
continuous.
%
Similar principles apply to category theory itself: a composition of
functors is functorial, a composition of natural transformations is
natural, etc.
%
In the case of category theory, these principles can be formalized
using algebraic structures such as 2-categories, bicategories,
(virtual) double categories and pro-arrow equipments\cite{}.
%
This approach has the advantage of avoiding the need to prove
the side conditions of naturality or enrichment.
%
A downside is that these algebraic structures are quite complicated,
and practitioners typically employ either an algebraic combinator
syntax or a 2-dimensional diagrammatic language that can be quite
beautiful and elegant, but is also seems very far removed from the
traditional formulation of category theory in terms of sets and
functions.

In this work, we apply the technique of categorical logic to these
algebraic structures to design a formal logic, which we call
\emph{virtual equipment type theory}, that can be soundly interpreted
in virtual equipments.
%
In \vett{}, the sorts are interpreted as categories, functions as
functors, relations as profunctors and implications of relations as
profunctor homomorphisms, a generalization of natural transformations.
%
The type theoretic syntax allows us to write all of our constructions
as if they were ordinary functors and profunctors in set-based
category theory, but the proofs are valid in an arbitrary virtual
equipments and so are valid in enriched and internal category theory
as well.
%
This is analogous to the situation in intuitionistic type theory,
where the syntax is similar to that of informal set theory, but by
restricting to intuitionistic reasoning, the constructions can all be
interpreted in models where types are spaces and terms are continuous
functions ``for free''.
%
The syntax provides an alternative to algebraic and
string-diagrammatic syntaxes for working with virtual equipments,
similar to how the lambda calculus provides an alternative to
categorical combinators and string diagram calculi for cartesian
closed categories.

\vett{} is a kind of \emph{ordered}, \emph{constructive} variant of
predicate logic over a unary type theory.
%
It is \emph{ordered} in that our notion of predicate is that of a
binary relation $R(\alpha;\beta)$ where the $\alpha$ is thought of as
a contravariant dependence and the $\beta$ has a covariant dependence,
and the rules of the logic distinguish between these two variances.
%
Proofs in the predicate logic $R_1, R_2,\ldots \vdash S$ are
restricted as in ordered linear logic \cite{lambek-calculus, ordered-linear-logic} in
that each input must be used exactly once and in the order they were
declared.
%
This order discipline ensures that every proof in the logic can be
interpreted as a \emph{natural} transformation.
%
The logic is \emph{constructive} in that we care about the identity of
proofs: they are natural transformations and there may be more than
one natural transformation of the same type.
%
If we ignore the identity of proofs, ordered predicate logic can be
viewed as a logic of preordered sets rather than categories.

The restriction to unary type theory and binary relations is fairly
strict: we cannot discuss important concepts such as functor
categories, opposite categories and arbitrary (co)-ends of
profunctors. However, the restriction is natural in that it
corresponds to a well-understood notion of model (virtual equipments)
and so we leave the extension to future work.
%
It is well known that a great deal of fundamental results and
constructions in category theory such as Yoneda and co-Yoneda lemmas,
adjoint functors and Kan extensions are already expressible in virtual
equipments, for instance virtual equipments are used for this purpose
in a text on $\infty$-category theory to avoid explicit manipulation
of model categories \cite{elements-of-oo-category-theory}.
%
We demonstrate some of these standard constructions
in \ref{sec:synthetic} to show that \vett{} gives a syntax in which
these formal constructions look like standard set-theoretic arguments.

Due to the restriction to unary functors and binary profunctors, it is
difficult to state some theorems internally to the theory. For
instance, since we cannot express functor categories, we cannot
express any theorem that quantifies over functors purely in ordered
predicate logic, and instead only as a metatheorem. To more easily
formalize these metatheorems, we add a third layer to the type theory:
an ordinary dependent type theory in the style of Martin-L\"of type
theory. Then all of our ordered predicate logic judgments will also be
indexed by a context from this dependent type theory, and the type
theory includes types for categories, functors, profunctors and
natural transformations that allow us to formalize theorems the logic
is too restrictive to encode. This is analogous to the notion of a
\emph{2-level type theory}: we add a dependent type theory for
reasoning about the terms of our virtual equipment type theory.

The dependent type theory layer also has an additional purpose: since
we can quantify over the structures in \vett{}, we can additionally
encode \emph{signatures} for \vett{} as simply contexts of the
dependent type theory. This allows for a clean formalization of many
categorical structures such as monoidal categories, or strong monads.

@@

\subsection{Basics of Profunctors}

While we assume the reader has some background knowledge of category
theory, we briefly define profunctors, which are not included in many
introductory texts. First, a category $\cat C$ has a collection of
objects and morphisms with identity and composition, and a functor $F
: \cat C \to \cat D$ is a function on objects and a function on
morphisms that respects source and target. It will be useful for
intuition to think of a category as a generalization of a preordered
set, which has a set of elements and a binary \emph{relation} on its
objects satisfying reflexivity and transitivity. It is fruitful to
think of a category as a \emph{proof-relevant preorder}, where
morphisms are thought of as proofs of ordering and the reflexivity and
transitivity constructions must additionally satisfy equations that
make them act like identity and composition. Then a functor is a
constructive version of a monotone function.

Given categories $\mathcal C$ and $\mathcal D$, a profunctor $R$ from
$\mathcal C$ to $\mathcal D$, written $R : \cat C \pto \cat D$ is a
functor $R : {\cat C}^{op} \times {\cat D} \to \Set$. We can think of
each set $R(c,d)$ as a set of ``heteromorphisms'' from $c$, an object
of $\cat C$ to $d$, an object of $\cat D$. Functoriality then defines
a notion of composition of heteromorphisms with homomorphisms in $\cat
C$ and $\cat D$. Thinking of categories as constructive preorders, a
profunctor is a constructive relation that is downward-closed in $\cat
C$ and upward-closed in $\cat D$, a kind of ``hetero-ordering''.
%
Given profunctors $R, S : \cat C \pto \cat D$, a homomorphism from
$R$ to $S$ is simply a natural transformation, which in the preordered
setting is simply an implication of relations.

The simplest example of a profunctor is for any category $\cat C$,
$\text{Hom}_{\cat C} : \cat C \pto \cat C$ is a profunctor. On
preorders this is just the ordering itself. The hom profunctor is the
unit for a composition of profunctors $R \odot S$ which is defined as
a \emph{co-end}. The heteromorphisms of $(R \odot S)(c,e)$ are triples
$(d \in \cat D, f \in R(c,d), g \in S(d,e))$ modulo the equivalence
relation generated by $(d,f,g) ~ (d', f', g')$ if there exists $h \in
D(d,d')$ such that $h\circ f = f'$ and $g'\circ h = g$. The
composition of profunctors is a generalization of the composition of
relations, and the hom profunctor is analogous in the same way to the
diagonal, or equality relation in that it is the identity for this
composition.
For any profunctor $R : \cat C \pto \cat D$ and functors $F : \cat
C' \to \cat C$ and $G : \cat D' \to \cat D$, by composition we can
define a profunctor $R(F,G)$, the ``restriction'' of $R$ along $F$ and
$G$.

Profunctors are not explicitly defined in most introductory texts on
category theory but arise implicitly in formulations of universal
properties such a representable functors and adjunctions.

\subsection{Paper Overview}

\begin{enumerate}
\item In \cref{sec:syntax} we introduce the syntax of Simple Category
  Theory.

@@

\item In \cref{sec:semantics}, we develop some basic model theory for
  ordered predicate logic, including a sound and complete notion of
  categorical model and sound interpretation in virtual equipments
  modeling ordinary, enriched and internal category theory.

>>>>

\begin{abstract}
  We present a domain-specific type theory for constructions and
  proofs in category theory. The type theory axiomatizes notions of
  category, functor, profunctor and a generalized form of natural
  transformations. The type theory imposes an ordered linear
  restriction on standard predicate logic, which guarantees that all functions
  between categories are functorial, all relations are profunctorial,
  and all transformations are natural by construction, with no
  separate proofs necessary. Important category theoretic proofs such
  as the Yoneda lemma and Co-yoneda lemma become simple type theoretic
  proofs about the relationship between unit, tensor and (ordered)
  function types, and can be seen to be ordered refinements of
  theorems in predicate logic.

@@

Category theory is a branch of mathematics that gives a systematic
theory of typed, ``higher-dimensional'' algebraic structures.
%
Originally developed for applications to homological algebra, it was
quickly discovered that (among many other applications) these
categorical structures were common in logic and computer science, and
that formal systems like logics, type theories and programming
languages can be soundly and completely modeled
by notions of structured categories\cite{lawvere,goguen,lambek,moggi}, a
phenomenon sometimes referred to as the Curry-Howard-Lambek
nncorrespondence \cite{curry-howard-lambek}.
%
Examples include the simply typed lambda calculus \cite{lambekscott},
computational lambda calculus \cite{moggi}, linear logic \cite{girard}
and dependent type theory\cite{not-sure}.
%
These seminal works have led to a fruitful field of research known as
categorical logic, where new type theories are designed based directly
on category-theoretic semantics.
%
The slogan is that the syntax of a type theory should present an
initial object in its category of models, a category-theoretic
reformulation and generalization of the logical notion of soundness
and completeness.

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex, and there is difficulty in
acquisition of competence in category theory by computer scientists
who do not have a strong background in pure mathemantics.
%
While much of this can be explained by lack of, or lack of awareness
of, accessible material for computer scientists, in some ways category
theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and Kan extensions produce a proliferation of ``naturality''
and ``functoriality'' side-conditions that must be discharged.
%
This is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
These generalizations have long been recognized as being important in
programming language applications: domain-, metric- and
step-index-enriched categories have been used to model general
recursion and internal categories have been used to model
parametricity and gradual
typing\cite{order-enriched-cats,topos-of-trees,reflexive-graphs,double-cats-gradual-typing},
to name a few.

% TODO: Yoneda example
Novices in category theory are often overwhelmed by the number of
naturality and functoriality side conditions that abound in otherwise
simple proofs in category theory. For instance consider the proof of the
Yoneda lemma, often the first ``big'' proof in an introductory category
theory text. The Yoneda lemma states that for any category $C$, functor
$P : C^{o} \to \Set$ (where $C^{o}$ is the ``opposite'' of $C$, i.e. 
$C$ with the direction of morphisms reversed) and object $A \in C$, the
set of natural transformations $C(-,A) \to P$ is in bijection with the
elements $P(A)$. Furthermore, this bijection is natural in both $A$ and
$P$. In the authors' experience teaching category theory the sheer
amount of naturality side conditions in this result can be very
intimidating to learners, while experienced practitioners will observe
that these naturality conditions are tedious to carry out but
routine. In the type theory we introduce in this paper, none of these
naturality conditions need to be explicitly proven. Instead, all terms
expressible in the type theory are automatically natural by
construction.

Fortunately, the tools of category theory itself can be employed to
simplify these constructions, specifically the tools of \emph{higher}
category theory.
%
Consider the analogous situation in differential calculus. When an
adept analyst writes down a function, they do not expand out the
$\epsilon\mathord{-}\delta$ definition of continuity and prove that it is
continuous. Rather, they are aware of certain \emph{syntactic
principles}, most importantly that composition of continuous functions
is continuous and that previously defined functions are known to be
continuous.
%
Similar principles apply to category theory itself: a composition of
functors is functorial, a composition of natural transformations is
natural, etc.
%
In the case of category theory, these principles can be formalized
using algebraic structures such as 2-categories, bicategories,
(virtual) double categories and pro-arrow equipments\cite{}.
%
This approach has the advantage of avoiding the need to prove
the side conditions of naturality or enrichment.
%
A downside is that these algebraic structures are quite complicated,
and practitioners typically employ either an algebraic combinator
syntax or a 2-dimensional diagrammatic language that can be quite
beautiful and elegant, but is also seems very far removed from the
traditional formulation of category theory in terms of sets and
functions.

In this work, we apply the technique of categorical logic to these
algebraic structures to design a formal logic, which we call
\emph{virtual equipment type theory}, that can be soundly interpreted
in virtual equipments.
%
In the ``standard model'' of \vett{}, the sorts are interpreted as
categories, functions as functors, relations as profunctors and
implications of relations as profunctor homomorphisms, a generalization
of natural transformations.
%
The type theoretic syntax allows us to write all of our constructions
as if they were ordinary functors and profunctors in set-based
category theory.
%
\vett also can be modeled in an arbitrary virtual equipments, and so the
constructions are valid in enriched and internal category theory as
well.
%%
%% DRL: I know what you mean, but I think conflating set theory and
%% type theory will put some people off... 
%% 
%% This is analogous to the situation in intuitionistic type theory,
%% where the syntax is similar to that of informal set theory, but by
%% restricting to intuitionistic reasoning, the constructions can all be
%% interpreted in models where types are spaces and terms are continuous
%% functions ``for free''.
%
The syntax provides an alternative to algebraic and
string-diagrammatic syntaxes for working with virtual equipments,
similar to how the lambda calculus provides an alternative to
categorical combinators and string diagram calculi for cartesian
closed categories.

\vett{} is an \emph{ordered}, \emph{constructive} variant of
predicate logic over a unary type theory.
%
It is \emph{ordered} in that our notion of predicate is that of a
binary relation $R(\alpha;\beta)$ where the $\alpha$ is thought of as
a contravariant dependence and the $\beta$ has a covariant dependence,
and the rules of the logic distinguish between these two variances.
%
Proofs in the predicate logic $R_1, R_2,\ldots \vdash S$ are
restricted as in ordered linear logic \cite{lambek-calculus, ordered-linear-logic} in
that each input must be used exactly once and in the order they were
declared.
%
This order discipline ensures that every proof in the logic can be
interpreted as a \emph{natural} transformation.
%
The logic is \emph{constructive} in that we care about the identity of
proofs: they are natural transformations and there may be more than
one natural transformation of the same type.
%
If we ignore the identity of proofs, ordered predicate logic can be
viewed as a logic of preordered sets rather than categories.

The restriction to unary type theory and binary relations is fairly
strict: we cannot discuss important concepts such as functor
categories, opposite categories and arbitrary (co)-ends of
profunctors. However, the restriction is natural in that it
corresponds to a well-understood notion of model (virtual equipments)
and so we leave these extensions to future work.
%
It is well known that a great deal of fundamental results and
constructions in category theory such as Yoneda and co-Yoneda lemmas,
adjoint functors and Kan extensions are already expressible in virtual
equipments, for instance virtual equipments are used for this purpose
in a text on $\infty$-category theory to avoid explicit manipulation
of model categories \cite{elements-of-oo-category-theory}.
%
We demonstrate some of these standard constructions
in \ref{sec:synthetic} to show that \vett{} gives a syntax in which
these formal constructions look like standard set-theoretic arguments.

However, we can work around these unary/binary restrictions to some
extent by viewing the type theory as a domain-specific language embedded
in a metalanguage.  For example, while we cannot talk about functor
categories inside the ordered part of the type theory, we can state a
theorem that quantifies over functors using the meta-language's
universal quantifier.  To more easily formalize these metatheorems, we
add a third layer to the type theory: an ordinary dependent type theory
in the style of Martin-L\"of type theory. Then all of our ordered
predicate logic judgments will also be indexed by a context from this
dependent type theory, and the type theory includes types for
categories, functors, profunctors and natural transformations that allow
us to formalize theorems the logic is too restrictive to encode. This is
analogous to the notion of a \emph{2-level type
  theory}\cite{voevodsky12,altenkirchXX}: we add a dependent type theory
for reasoning about the terms of our virtual equipment type theory.
The dependent type theory layer also has an additional purpose: since
we can quantify over the structures in \vett{}, we can additionally
encode \emph{signatures} for \vett{} as simply contexts of the
dependent type theory. This allows for a clean formalization of many
categorical structures such as monoidal categories, or strong monads.

@@

\paragraph{Basics of Profunctors.}

While we assume the reader has some background knowledge of category
theory, we briefly define profunctors, which are not included in many
introductory texts.  Profunctors arise implicitly in formulations of
universal properties such as representable functors and adjunctions.

First, a category $\cat C$ has a collection of objects and morphisms
with identity and composition, and a functor $F : \cat C \to \cat D$ is
a function on objects and a function on morphisms that respects source
and target and preserves identity and composition. It will be useful for
intuition to think of a category as a generalization of a preordered
set, which has a set of elements and a binary \emph{relation} on its
objects satisfying reflexivity and transitivity. It is fruitful to think
of a category as a \emph{proof-relevant preorder}, where morphisms are
the proofs of ordering, and the reflexivity and transitivity
constructions must additionally satisfy equations that make them act
like identity and composition. Then a functor is a proof-relevant
version of a monotone function.

Given categories $\mathcal C$ and $\mathcal D$, a profunctor $R$ from
$\mathcal C$ to $\mathcal D$, written $R : \cat C \pto \cat D$ is a
functor $R : {\cat C}^{op} \times {\cat D} \to \Set$. We can think of
each set $R(c,d)$ as a set of ``heteromorphisms'' from $c$ (an object of
$\cat C$) to $d$ (an object of $\cat D$).  These ``heteromorphisms''
differ from the usual ``homomorphisms'' in a single category $\cat C$
because they relate two objects from two different categories.
Functoriality then defines a notion of composition of heteromorphisms
with homomorphisms in $\cat C$ and $\cat D$. Thinking of categories as
proof-relevant preorders, a profunctor is a proof-relevant relation that
is downward-closed in $\cat C$ and upward-closed in $\cat D$, a kind of
``hetero-ordering''.
%
Given profunctors $R, S : \cat C \pto \cat D$, a homomorphism from
$R$ to $S$ is simply a natural transformation, which in the preordered
setting is simply an implication of relations.

In addition to their usefulness for formalizing category theory, one
reason we take profunctors to be a basic concept of \vett{} is that they
allow us to give a \emph{universal property} for the type of ``morphisms
in a category ${\cat C}$''.  This is analogous to how the $J$ elimination
rule for the identity type in Martin-L\"of type theory gives a universal
property for morphisms in a groupoid (the special case of a category
where all morphisms are
invertible)~\cite{hofmannstreicher,awodey,voevodsky}. The reason
profunctors are useful for this purpose is that, for any category $\cat
C$, $\text{Hom}_{\cat C} : \cat C \pto \cat C$ is a profunctor. On
preorders this is just the preorder's ordering relation
itself. Moreover, the hom profunctor is the unit for a composition of
profunctors $R \odot S$ which is defined as a \emph{co-end}. The
heteromorphisms of $(R \odot S)(c,e)$ are triples $(d \in \cat D, f \in
R(c,d), g \in S(d,e))$ modulo the equivalence relation generated by
$(d,f,g) ~ (d', f', g')$ if there exists $h \in D(d,d')$ such that
$h\circ f = f'$ and $g'\circ h = g$. The composition of profunctors is a
generalization of the composition of relations, and just as the equality
relation is the identity for the composition of relations, the hom
profunctor is the identity for this composition.  The unit law for the
hom profunctor can be seen as a ``morphism induction'' principle,
analogous to the ``path induction'' used in homotopy type theory (though
in this paper we consider only ordinary 1-dimensional categories, not
higher generalizations).

%% orphaned --- move later? 
%% For any profunctor $R : \cat C \pto \cat D$ and functors $F : \cat C'
%% \to \cat C$ and $G : \cat D' \to \cat D$, by composition we can define a
%% profunctor $R(F,G)$, the ``restriction'' of $R$ along $F$ and $G$.

\paragraph{Paper Overview.}

\begin{enumerate}
\item In \cref{sec:syntax} we introduce the syntax of \vett{}.

@@

\item In \cref{sec:semantics}, we develop some basic model theory for
  \vett{}, including a sound and complete notion of categorical model
  and sound interpretation in virtual equipments modeling ordinary,
  enriched and internal category theory.
